# homework3: plot

## Load library

```{r}
library(sf)
library(here)
library(janitor)
library(tidyverse)
library(terra)
library(ggplot2)
```

## Read in data

```{r}
spain <- sf::st_read(here("data",
                          "wk3", 
                          "gadm41_ESP.gpkg"),
                      layer='ADM_ADM_0')

layers <- st_layers(here("data",
                         "wk3", 
                         "gadm41_ESP.gpkg"))

world_cities <- sf::st_read(here("data", 
                                 "wk3", 
                                 "World_Cities", 
                                 "World_Cities.shp"))

ssp1 <-terra::rast(here("data", 
                        "wk3",
                        "wc2.1_2.5m_tmax_ACCESS-CM2_ssp126_2081-2100.tif"))

ssp1_mean <- mean(ssp1)

ssp5 <-terra::rast(here("data", 
                        "wk3", 
                        "wc2.1_2.5m_tmax_ACCESS-CM2_ssp585_2081-2100.tif"))
```

## Filter cities

```{r}
spanish_cities <- world_cities %>%
  janitor::clean_names()%>%
  dplyr::filter(cntry_name=="Spain")

#ssp_diff <- ssp5-ssp1
```

## Crop and mask rasters

```{r}
spain_diff <- ssp1 %>%
  terra::crop(.,spain)
exact_spain <- spain_diff %>%
  terra::mask(.,spain)

spain_diff5 <- ssp5 %>%
  terra::crop(.,spain)
exact_spain5 <- spain_diff5 %>%
  terra::mask(.,spain)
```

## Subtract rasters

```{r}
diff_climate_model <- exact_spain5 - exact_spain 
```

## Rename tiff stack and extract data from points

```{r}
month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
names(diff_climate_model) <- month
```

## Extract spain point

```{r}
spain_city_diff<- terra::extract(diff_climate_model, spanish_cities)
```

## Join the extracted data back to the point dataset...either this...

```{r}
spain_city_diff2 <- spanish_cities %>% 
  add_column(spain_city_diffJan=spain_city_diff[,2])%>%
  add_column(spain_city_diffFeb=spain_city_diff[,3])%>%
  add_column(spain_city_diffMar=spain_city_diff[,4])%>%
  add_column(spain_city_diffApril=spain_city_diff[,5])%>%
  add_column(spain_city_diffMay=spain_city_diff[,6])%>%
  add_column(spain_city_diffJune=spain_city_diff[,7])%>%
  add_column(spain_city_diffJuly=spain_city_diff[,8])%>%
  add_column(spain_city_diffAug=spain_city_diff[,9])%>%
  add_column(spain_city_diffSept=spain_city_diff[,10])%>%
  add_column(spain_city_diffOct=spain_city_diff[,11])%>%
  add_column(spain_city_diffNov=spain_city_diff[,12])%>%
  add_column(spain_city_diffDec=spain_city_diff[,13])

```

## Or make a join ID column in the original point sf and use that to join...

```{r}
spanish_cities_join_ID <- spanish_cities %>%
  dplyr::mutate(join_id= 1:n()) # 为每一行创建一个标识列。n()是数据框行数。
```

## Now join

```{r}
spain_city_diff2 <- spanish_cities_join_ID%>%
  dplyr::left_join(., # ...ID放在...diff文件的左边。
            spain_city_diff,
             by = c("join_id" = "ID"))
```

## Drop the geometry and make a facet plot of differences per month

```{r}
# 取月份值，去掉几何信息
city_climate_diff <- spain_city_diff2 %>%
  dplyr::select(c(,16:27))%>%
  sf::st_drop_geometry(.)%>%
  dplyr::as_tibble()

# pivot
tidy_city_diff <- city_climate_diff %>%
  tidyr::pivot_longer(everything(), 
               names_to="Months", 
               values_to="temp_diff")

# 把月份变成factor变量，确保输出顺序是按月份，而不是按首字母。
facet_plot <- tidy_city_diff %>%
  dplyr::mutate(Months = factor(Months, levels = c("Jan","Feb","Mar",
                                                   "Apr","May","Jun",
                                                   "Jul","Aug","Sep",
                                                   "Oct","Nov","Dec")))

# Plot faceted histogram
plot <- ggplot(facet_plot, aes(x=temp_diff, na.rm=TRUE)) +  # 根据facet_plot创建ggplot对象，映射temp_diff到x轴，忽略NA值
  geom_histogram(color="black", binwidth=.1) +  # 绘制直方图，黑色边框，柱宽为0.1
  labs(title="Ggplot2 faceted difference in climate scenarios of max temp",  # 标题
       x="Temperature",  # x轴标签为“Temperature”
       y="Frequency") +  # y轴标签为“Frequency”
  facet_grid(Months ~ .) +  # 按月份（Months）分面
  theme(plot.title = element_text(hjust = 0.5))  # 标题居中

plot

```

## Extra - violin and box plot

How about taking each climate scenario and plotting the distribution of max temp values for Spanish cities....

```{r}

####### extra analysis####when i assumed it was just a single year not months

ssp1 <-terra::rast(here("data", "wk3", "wc2.1_2.5m_tmax_ACCESS-CM2_ssp126_2081-2100.tif"))
ssp2 <-terra::rast(here("data", "wk3", "wc2.1_2.5m_tmax_ACCESS-CM2_ssp245_2081-2100.tif"))
ssp3 <-terra::rast(here("data", "wk3", "wc2.1_2.5m_tmax_ACCESS-CM2_ssp370_2081-2100.tif"))
ssp5 <-terra::rast(here("data", "wk3", "wc2.1_2.5m_tmax_ACCESS-CM2_ssp585_2081-2100.tif"))

ssp1mean <- ssp1 %>%
  terra::crop(.,spain)%>%
  terra::mask(.,spain)%>%
  mean(.)

ssp2mean <- ssp2 %>%
  terra::crop(.,spain)%>%
  terra::mask(.,spain)%>%
  mean(.)
  
ssp3mean <- ssp3 %>%
  terra::crop(.,spain)%>%
  terra::mask(.,spain)%>%
  mean(.)
  
ssp5mean <- ssp5 %>%
  terra::crop(.,spain)%>%
  terra::mask(.,spain)%>%
  mean(.)
  
spain_city_ssp1<- terra::extract(ssp1mean, spanish_cities)
spain_city_ssp2<- terra::extract(ssp2mean, spanish_cities)
spain_city_ssp3<- terra::extract(ssp3mean, spanish_cities)
spain_city_ssp5<- terra::extract(ssp5mean, spanish_cities)

spain_city3 <- spanish_cities %>% 
  add_column(ssp1=spain_city_ssp1$mean)%>%
  add_column(ssp2=spain_city_ssp2$mean)%>%
  add_column(ssp3=spain_city_ssp3$mean)%>%
  add_column(ssp5=spain_city_ssp5$mean)

spain_city4 <- spain_city3 %>%
  as_tibble()%>%
  dplyr::select(city_name, ssp1, ssp2, ssp3, ssp5)%>%
  pivot_longer(cols = 2:5,
              names_to = "model",
              values_to = "max_temp"
  )

violin <- ggplot(spain_city4, 
                 aes(x=model, y=max_temp, fill=model)) + # fill=name allow to automatically dedicate a color for each group
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75))+
  geom_point()
#geom_jitter(height = 0, width = 0.1)

box <- ggplot(spain_city4, aes(x=model, y=max_temp, fill=model)) + # fill=name allow to automatically dedicate a color for each group
  geom_boxplot()+
  theme_minimal()+
  labs(
    x="Model", 
    y="Climate projection max tempearture")+
  theme(legend.position = "none")

violin
box

# we could also create a violin plot for the whole of spain
# per model or use metro areas. 
```
