# homework2: calculate and plot

## Task

calculate the average percent of science students (in all) grades per county meeting the required standards

We have:

-   Count met standard
-   Count expected to test
-   test subject

Need to filter for science and mutate to work out % that have met

## Load library

```{r}
library(tidyverse)
library(here)
library(sf)
library(janitor) # 数据清理，如清理列名。
library(tmap)
```

## Load data

-   `na='NULL'`: 将原表中的'NULL'都设为na(R读得懂的空值)

```{r}
shape <- st_read(here::here("data", 
                            "wk2",            "Washington_Counties_with_Natural_Shoreline___washsh_area",                           "Washington_Counties_with_Natural_Shoreline___washsh_area.shp"))

report <- read_csv(here::here("data", 
                              "wk2", 
 "Report_Card_Assessment_Data_2018-19_School_Year_20241110.csv"),
                              na= "NULL") 
```

## check data type (create a new dataframe to show them)

-   `summarise_all(class)`: 返回所有列的type
-   `pivot_longer(everything())`: 所有列转换成长格式
-   `names_to='<name>'`: 原始列名放进新列'name'中
-   `values_to='<name>'`: 原始数据类型放进新列'name'中

```{r}
Datatypelist <- report %>% 
  summarise_all(class) %>% 
  pivot_longer(everything(), 
               names_to="All_variables",
               values_to="Variable_class")

Datatypelist
```

## clean and calculate (county)

-   `clean_names()`: 标准列名清理，主要是变小写，加下划线。比如OrgLevel会变成org_level
-   `select()`: 选择需要的列
-   `filter(<column> == 'x')`: 筛选
-   `filter(<column> != 'x')`: 筛选，某列除了x以外所有的
-   `group_by(<column>)`: 按某列分组
-   `na.omit()`: 删除包含空值的行
-   `summarise()`: 计算，和前面的group_by()联用，分组计算，比如分别算每个county的数。结果是每个county对应一行。如果不用summarise，也能计算每个county的数，但是行数不会减少（一个county对应很多行），会有重复行。
-   `mutate()`: 计算并添加列

```{r}
county_only <- report %>%
  clean_names() %>%
  select(county, organization_level, test_subject, count_met_standard, 
         count_of_students_expected_to_test, grade_level)%>%
  filter(county != "Multiple")%>%
  filter(organization_level == "School")%>%
  filter(test_subject == "Science")%>%
  filter(grade_level=="All Grades")%>%
  group_by(county)%>%
  na.omit()%>%
  
  summarise(total_county_met_standard=sum(count_met_standard), 
            total_county_to_test=sum(count_of_students_expected_to_test))%>%
  mutate(percent_met_per_county=(total_county_met_standard/total_county_to_test)*100)
```

## calculate for the whole state

-   `pull()`: 从数据框中提取单列。这里只有一列，就直接返回一个value

```{r}
state_average <- mean(county_only$percent_met_per_county)

#or 
state_average2 <- county_only%>%
  summarise(state_average= mean(percent_met_per_county))%>%
  pull()

#or
state_average3 <- county_only%>%
  select(percent_met_per_county)%>%
  summarise(state_average=mean(percent_met_per_county))%>%
  pull()
```

## compare county average to the state average, above or below?

-   `mutate(across(<column>, round, 0))`: 四舍五入，保留0位小数
-   `case_when(<condition> ~ 'x')`: 条件语句，condition时赋值x

```{r}
county_only_above_below_state <- county_only %>%
  mutate(difference_to_state=(percent_met_per_county-state_average))%>%
  mutate(across(difference_to_state , round, 0))%>%
  mutate(above_below = case_when(difference_to_state<0 ~ "below",
                                   difference_to_state>0 ~ "above",
                                   difference_to_state==0 ~ "equal"
                                   ))
```

## Join to spatial data

-   `clean_names(.)`: 点号表示作用于管道前的第一个变量，这里也可以不加。
-   `left_join()`: join两个数据框，以左为准，把右边有数据的加过来。
-   `by = c('<column1> = '<column2>')`: 根据这两个列来join。c用来选择多个列。

```{r}
joined_data <- shape %>% 
  clean_names(.) %>%
  left_join(., 
            county_only_above_below_state,
            by = c("countylabe" = "county"))
```

## plot

```{r}
#bbox_county <- joined_data %>%
#  st_bbox(.) %>% 
#  tmaptools::read_osm(., type = "osm", zoom = NULL)

#tm_shape(bbox_county)+
#  tm_rgb()+

  tm_shape(joined_data) + 
  tm_polygons("above_below", 
             # style="pretty",
              palette="Blues",
              midpoint=NA,
              #title="Number of years",
              alpha = 0.5) + 
  tm_compass(position = c("left", "bottom"),type = "arrow") + 
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(title = "Counties above or below state avearge for science in all grades", 
            legend.position = c("right", "bottom"))
```
